---
title: "07_analysis_3"
author: Antoine Andréoletti (s242830), Olivier Gaufrès (s243252), Amy Surry (s201902), Lea Skytthe (s203531), Trine Søgaard (s204655)
format:
  html:
    embed-resources: true
editor: visual
---

# Differential Expression Analysis

```{r}
#| label: 07_load-libraries
#| message: False
#| warning: False

rm(list=ls())

library("tidyverse")
library("pheatmap")
library("ggrepel")
library("DESeq2") # DEA libraries

source(file.path("99_proj_func.R"))
```

## Load data

```{r}
#| label: 07_load-data

data_pheno <- read.csv("../data/03_metadata_aug.csv")

# We want patients as row names, so we convert the tibble to a dataframe and names the rows
data_pheno <- data.frame(data_pheno) |> 
  column_to_rownames("source_name")

data_counts.total <- read.csv("../data/03_count_data_wide_aug.csv")

# We want genes as row names, so we convert the tibble to a dataframe and names the rows
data_counts <- data.frame(data_counts.total) |>
  dplyr::select(-gene_id, -gene_symbol, -transcript_ID)

colnames(data_counts) <- colnames(data_counts) |> 
  substring(2)

# The output of the mapping algorithm they use is an estimated number of reads mapped to each gene. It is a float while we need an integer. We will then convert it.
data_counts <- data_counts |> 
  mutate(across(where(is.numeric), as.integer))
```

## Differential Expression Analysis

### Using DESeq2

```{r}
#| label: 07_dea-deseq2-general
#| warning: false
#| message: false

dds <- DESeqDataSetFromMatrix(countData = data_counts,
                              colData = data_pheno,
                              design = ~disease)

featureData <- data.frame(gene=rownames(data_counts))
mcols(dds) <- DataFrame(mcols(dds), 
                        featureData)

# Adding the reference level (normal)
dds$disease <- relevel(dds$disease, 
                       ref = "normal")

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, 
            by = "transcript_ID")

# Creating the folder to save the data

DESeq2_data_dir <- "../data/DESeq2"
if( !dir.exists(DESeq2_data_dir) ){
  dir.create(path = DESeq2_data_dir)
}

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-global.csv"))
```

### Volcano plots

```{r}
#| label: 07_plot-volcano
#| warning: false
#| message: false

# General analysis
plot_volcano(x = res, 
             title = "general",
             title_plot = "sick against healthy",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = data_pheno |> select(disease),
                  title = "global",
                  title_plot = "Global"
                  )
```

## Testing now in fasting conditions

```{r}
#| label: 07_dea-deseq2-fasting
#| warning: false
#| message: false

data_pheno_fasting <- data_pheno |> 
  dplyr::filter(clinical_information == "fasting")

data_counts_fasting <- data_counts |> 
  dplyr::select(rownames(data_pheno_fasting))

dds <- DESeqDataSetFromMatrix(countData = data_counts_fasting,
                              colData = data_pheno_fasting,
                              design = ~disease)

featureData <- data.frame(gene=rownames(data_counts_fasting))
mcols(dds) <- DataFrame(mcols(dds), 
                        featureData)

# Adding the reference level (normal)
dds$disease <- relevel(dds$disease, ref = "normal")

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, 
            by = "transcript_ID")

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-fasting.csv"))

plot_volcano(x = res, 
             title = "fasting", 
             title_plot = "in fasting conditions, sick against healthy",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = data_pheno_fasting |> select(clinical_information),
                  title = "fasting",
                  title_plot = "Fasting conditions - Sick vs healthy"
                  )
```

## Testing now in postprandial condition

```{r}
#| label: 07_dea-deseq2-postprandial
#| warning: false
#| message: false

data_pheno_postprandial <- data_pheno |> 
  dplyr::filter(clinical_information == "postprandial")

data_counts_postprandial <- data_counts |> 
  dplyr::select(rownames(data_pheno_postprandial))

dds <- DESeqDataSetFromMatrix(countData = data_counts_postprandial,
                              colData = data_pheno_postprandial,
                              design = ~disease)

featureData <- data.frame(gene=rownames(data_counts_postprandial))
mcols(dds) <- DataFrame(mcols(dds), 
                        featureData)

# Adding the reference level (normal)
dds$disease <- relevel(dds$disease, 
                       ref = "normal")

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, by = "transcript_ID")

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-postprandial.csv"))

# Plot the volcano
plot_volcano(x = res, 
             title = "postprandial", 
             title_plot = "in postprandial conditions, sick against healthy",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = data_pheno_postprandial |> select(disease),
                  title = "postprandial",
                  title_plot = "Postprandial conditions - Sick vs healthy"
                  )
```

## Difference in expression fasting or not cirrhosis

```{r}
#| label: 07_dea-deseq2-cirrhosis
#| warning: false
#| message: false

data_pheno_cirrhosis <- data_pheno |> 
  dplyr::filter(disease == "cirrhosis of liver")

data_counts_cirrhosis <- data_counts |> 
  dplyr::select(rownames(data_pheno_cirrhosis))

dds <- DESeqDataSetFromMatrix(countData = data_counts_cirrhosis,
                              colData = data_pheno_cirrhosis,
                              design = ~clinical_information)

featureData <- data.frame(gene=rownames(data_counts_cirrhosis))
mcols(dds) <- DataFrame(mcols(dds), 
                        featureData)

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, by = "transcript_ID")

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-cirrhosis.csv"))

plot_volcano(x = res, 
             title = "cirrhosis", 
             title_plot = "for cirrhosis patients, fasting against postprandial",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = data_pheno_cirrhosis |> select(clinical_information),
                  title = "cirrhosis",
                  title_plot = "Cirrhosis patients - Fasting vs postprandial"
                  )
```

## Difference in expression fasting or not non-alcoholic fatty liver disease

```{r}
#| label: 07_dea-deseq2-NAFL
#| warning: false
#| message: false

data_pheno_NAFL <- data_pheno |> 
  dplyr::filter(disease == "non-alcoholic fatty liver disease")

data_counts_NAFL <- data_counts |> 
  dplyr::select(rownames(data_pheno_NAFL))

dds <- DESeqDataSetFromMatrix(countData = data_counts_NAFL,
                              colData = data_pheno_NAFL,
                              design = ~clinical_information)

featureData <- data.frame(gene=rownames(data_counts_NAFL))
mcols(dds) <- DataFrame(mcols(dds), 
                        featureData)

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, by = "transcript_ID")

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-NAFL.csv"))

plot_volcano(x = res, 
             title = "NAFL", 
             title_plot = "for non-alcoholic patients, fasting against postprandial",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = data_pheno_fattyliver |> select(clinical_information),
                  title = "non-alcoholic-fatty-liver-disease",
                  title_plot = "Bon-alcoholic patients - Fasting vs postprandial"
                  )
```

## Difference in expression fasting or not in healthy people

```{r}
#| label: 07_dea-deseq2-normal
#| warning: false
#| message: false

data_pheno_normal <- data_pheno |> 
  dplyr::filter(disease == "normal")

data_counts_normal <- data_counts |> 
  dplyr::select(rownames(data_pheno_normal))

dds <- DESeqDataSetFromMatrix(countData = data_counts_normal,
                              colData = data_pheno_normal,
                              design = ~clinical_information)

featureData <- data.frame(gene=rownames(data_counts_normal))
mcols(dds) <- DataFrame(mcols(dds), featureData)

# DEA
dds <- DESeq(dds)
res <- results(dds)

gene_symbol.df <- data_counts.total |> 
  dplyr::select(gene_symbol) |> 
  rownames_to_column("transcript_ID")

res <- as.data.frame(res) |> 
  rownames_to_column("transcript_ID") |> 
  left_join(gene_symbol.df, by = "transcript_ID")

#Saving the results
write_csv(as.data.frame(res), 
          file.path("../data/DESeq2/03_DESeq2_results-normal.csv"))

plot_volcano(x = res, 
             title = "normal", 
             title_plot = "for healthy people, fasting against postprandial",
             fc_cutoff = log2(2), 
             p_cutoff = 0.05)

annotation_data_pheno <-  data_pheno_normal |> 
  dplyr::select(clinical_information)

plot_topN_heatmap(res = res,
                  dds = dds,
                  N = 10,
                  annotation_data_pheno = annotation_data_pheno,
                  title = "normal",
                  title_plot = "Healthy people - Fasting vs Postprandial"
                  )
```
