---
title: "Analysing metadata"
author: Antoine Andréoletti, Olivier Gaufrès, Amy Surry, Lea Skytthe, Trine Søgaard
format:
  html:
    embed-resources: true
editor: visual
---

The objective of this file is to analyse the patients we have.

## Importing libraries

```{r}
#| message: False

library("tidyverse")
library("table1")
```

## Importing the data

```{r}
metadata <- read_csv("../data/metadata.csv")
metadata |> sample_n(10)
```

```{r}
metadata |> 
  mutate(sex = factor(sex),
         disease = factor(disease)) |> 
  table1(x = formula(~ sex + age | disease),
         data = _)
```

We will now remove excluded people from our analysis.

```{r}
metadata <- metadata |> 
  filter(disease != "excluded")
```

## Analysing age, gender and disease distributions

```{r}
metadata |> 
  mutate(age_group = cut(
    x = age,
    breaks = seq(from = 0,
                 to = 100,
                 by = 10))) |> 
  ggplot(mapping = aes(x = age_group, fill = sex)) +
    geom_histogram(stat = "count", 
                   position = position_dodge(preserve = "single"))
```

```{r}
metadata |> 
  ggplot(mapping = aes(x = disease, fill = sex)) +
  geom_histogram(stat = "count",
                 position = position_dodge(preserve = "single")) 
```

```{r}
metadata |> 
  ggplot(mapping = aes(x = disease, y = age, fill = sex)) +
    geom_boxplot()
```

Lets make a statistical analysis to see if there is any significant difference in the average age of people with cirrhosis disease as it looks like in the above plot.

```{r}
male_cirrhosis <- metadata |> 
  filter(sex == "male", disease == "cirrhosis of liver") |> 
  pull(age)

female_cirrhosis <- metadata |> 
  filter(sex == "female", disease == "cirrhosis of liver") |> 
  pull(age)

t.test(male_cirrhosis, female_cirrhosis)
```

The p-value is above 0.05 so we can say that there is no significant difference between the two groups.
